---
title: "Homework 4"
author: "Gregory Bushman"
output: word_document
---
Homework 4: Scale Scores and Multiple Regression  

**Research Question and Hypothesis**  

The Flint Adolescent Study includes a number of psychometric scales that measure factors related to an individual's alcohol use, as well as the influences and pressures around them related to alcohol use behaviors. Through multiple regression, one can assess for the impact of things like future orientation to alcohol use, peer alcohol use, and close adult alcohol use on an individual's own, personal alcohol use. I predict that peer alcohol use and close adult alcohol use will be significant factors in predicting an individual's alcohol use because high levels of peer or close adult alcohol use may signal a normalization of alcohol use within an individual's social network. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

**Initialize and Import**

```{r initialize}
# load packages
library(tidyverse)
library(GGally)
library(psych)
library(car)

# import data
fas <- read.csv("C:/Users/Greg/Documents/Winter 2017/Quantitative Methods/fas_wave1.csv")
```

**Prepare for Analysis**  

```{r prep}
# filter data to required columns
scales <- fas %>%
  select(ageiw1, a2, a3, 
         i4a, i4b, i4c,          #alcohuse
         e3, e9a, e9b, g1h, d1j, #futrornt
         c2b:c2f,                #frnuseal
         f3e:f3m)                #adultdrg

# recode columns that need it: i4a, e3
scales$i4ar <- recode(fas$i4a, '1=5; 2=4; 3=3; 4=2; 5=1; else=NA', as.numeric = T)
scales$e3r  <- recode(fas$e3, '1=5; 2=4; 3=3; 4=2; 5=1; else=NA', as.numeric = T)

# make outcome columns numeric
scales <- data.frame(scales[1:3], lapply(scales[4:27], as.numeric))

# mutate new scale score columns
scales <- scales %>%
  mutate(
    alcohuse = rowSums(.[c("i4ar", "i4b", "i4c")], na.rm = T),
    futrornt = rowMeans(.[c("e3r", "e9a", "e9b", "g1h", "d1j")], na.rm = T),
    frnuseal = rowMeans(.[c("c2b", "c2c", "c2d", "c2d", "c2e", "c2f")], na.rm = T),
    adultdrg = rowMeans(.[c("f3e", "f3f", "f3g", "f3h", "f3i", 
                            "f3j", "f3k", "f3l", "f3m")], na.rm = T)
  )

scales$alcohuse <- ifelse(scales$alcohuse == 0, NA, scales$alcohuse)

# factor grouping variables
scales$a2 <- factor(scales$a2, levels = c(1, 2), labels = c("Male", "Female"))
scales$a3 <- factor(scales$a3, levels = c(1, 2, 3), labels = c("Black", "White", "Mixed"))
```

**Test Assumptions**  

From these tests, we can see that alcoholuse is relatively normally distributed, with a mean of 7.26 and a standard deviation of 1.98. Future orientation, is skewed to the right, while peer alcohol use and close adult alcohol use are skewed to the left. Given that one of the assumptions of regression analysis is the normality of the variables being used, this could have effects on the reliability of the regression analysis. Another assumption of regression is that variance between groups is about equal. Given that the levene test for race is significant, suggesting that the variance between racial groups is not equal, this sample may also violate the assumption of homoscedasticity. 

*Alcohol Use*  

```{r}
# alcohuse
describe(scales$alcohuse)
levene.test(scales$alcohuse, scales$a2)
levene.test(scales$alcohuse, scales$a3)
ggplot(scales, aes(alcohuse)) + geom_density() + theme_minimal()
```

*Future Orientation*

```{r}
# furtornt
describe(scales$futrornt)
ggplot(scales, aes(futrornt)) + geom_density() + theme_minimal()
```

*Peer Alcohol Use*

```{r}
# frnuseal
describe(scales$frnuseal)
ggplot(scales, aes(frnuseal)) + geom_density() + theme_minimal()
```

*Close Adult Drug Use*

```{r}
# adultdrg
describe(scales$adultdrg)
ggplot(scales, aes(adultdrg)) + geom_density() + theme_minimal()
```

**Multiple Regression**  

Here, six regression models have been fitted. Models 1-3 are models of alcohol use, explained by each of the three other variables of interest (future orientation, peer alcohol use, and close adult alcohol use). These models are also set up to control for age, sex, and race. Model four, which has the best adjusted R-squared value of the first four models (0.071), uses only peer and close adult alcohol use, sex, and race to predict an individual's alcohol use. Based on this model, only peer drug use was a significant predictors of alcohol use, with peer alcohol use having the largest impact on alcohol use. Given the levene's test earlier which showed that racial groups did not display equal variance with respect to alcohol use, I decided to subset this fourth model by two of the three racial groups. When model four was run after being subset to only look at black study participants (model 4.1), the adjusted R-squared of the model increased. In this subset model, both peer alcohol use and sex were significant predictors. When model four was subset to only look at white participants (model 4.2), no predictors were significant. 

```{r}
# Regression 1
reg.1 <- lm(alcohuse ~ futrornt + ageiw1 + a2 + a3, data = scales)
summary(reg.1)

# Regression 2
reg.2 <- lm(alcohuse ~ frnuseal + ageiw1 + a2 + a3, data = scales)
summary(reg.2)

# Regression 3
reg.3 <- lm(alcohuse ~ adultdrg + ageiw1 + a2 + a3, data = scales)
summary(reg.3)

# Regression 4
reg.4 <- lm(alcohuse ~ frnuseal + adultdrg + a2 + a3, data = scales)
summary(reg.4)

# Regression 4.1
reg.4.1 <- lm(alcohuse ~ frnuseal + adultdrg + a2, data = scales, subset = a3 == "Black")
summary(reg.4.1)
anova(reg.4.1)
confint(reg.4.1)

# Regression 4.2
reg.4.2 <- lm(alcohuse ~ frnuseal + adultdrg + a2, data = scales, subset = a3 == "White")
summary(reg.4.2)
```

**Diagnostics**  

These plots show that there is low auto-correlation between the variables of interest, and that the model fits well. The scale-location plot does indicate that there is not perfect homoscedasticity; however, this was something that was already known based on the levene's test performed earlier. 

```{r}
# plots
plot(reg.4.1)

# independence: durbin-watson test
dwt(reg.4.1)

# multicolinearity
ggpairs(scales[28:31])
```

**Conclusions**

The differences between models 4, 4.1, and 4.2 suggest that there are significant racial differences in how these factors of peer alcohol use, close adult alcohol use, and sex are interacting with an individual's alcohol use behaviors. Black women in this study seem to be especially vulnerable to peer pressure when it comes to making their own alcohol use decisions. 
